---
title: "MSB1015 - GSE270199 Processing"
author: "Anne"
date: "2026-10"
output: html_document
---
#Loading the data
Make sure that the previous file has been run and has saved the rds files (GSE270199_imputed.rds & GSE270199_redmeta.rds found in the OUTPUTS location)

#Clear the global environment
  Clear the environment to reduce the chance of issues on reruns
```{r clear_global}
rm(list=ls())
```
#Setting the working directory (wd)
Other files are assumed to be located in the same, or downstream maps from the location of this file. The location of this file is used as a reference point to find the other files. 
```{r working_directory}
#download and load the package for setting the working directory to the current file
if (!require("rstudioapi", quietly = TRUE))
  install.packages("rstudioapi")
library(rstudioapi)  

#Set the wd to the location of this file  
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```
#Library installations and loading 
As I will be using a project (renv) for the MSB1015 project, I check first that the packages that need to be installed are installed.
```{r library_R_file, include=FALSE}
#For cleanliness the checks, installing and loading of the various libraries is in a separate .R file. This file is assumed to be in the same map as this file (in the wd).
source("libraries.R")
```

```{r library_installation}
#Use the function to install the libraries
Fun_install_libraries()
```

```{r library_loading}
#Use the function to load the libraries
Fun_load_libraries()
```
#Load in the checks R file
There are some functions to run checks for various things in this separate .R file.
These are not in the main code as they create quite a bit of bloat that makes the code more difficult to read.
These are not reusable functions
```{r checks}
source("checks.R")
```
#Load in the custom_functions R file
There are some functions to run checks for various things in this separate .R file.
These are not in the main code to make the main script less bloated and the custom functions easier to find (instead of all of them being spread over the Rmd file).
```{r custom_functions}
source("custom_functions.R")
```
#File locations
Set the locations of the files that are loaded in and the locations on where the outputs will go. 
This code will also do a check that the given locations are findable. 
```{r file_locations}
#Make sure to save the wd so that it is easy to use to set file locations
working_directory <- getwd()

#The default INPUT_FILE_LOCATION is in a map where this file is also saved.
INPUT_FILE_LOCATION <- paste0(working_directory,"/INPUTS")
#The default OUTPUT_FILE_LOCATION is in a map where this file is also saved.
OUTPUT_FILE_LOCATION <- paste0(working_directory,"/OUTPUTS")

#-----#

#Make sure that the file locations can be found and in the case of the outputs create a dir if it doesn't exist
Fun_in_out_check()

```
#Settings
Any global settings that are used are set here.
*Please make sure the settings are the same between the files when applicable* 
```{r settings}
#-----#
#In multiple sections:
  #During testing it was often beneficial to run the whole file. Preventing the plots from being printed allows the file to run slightly faster.
  #This setting is mostly used for testing purposes. 
#TAG: add into the settings check
RUN_PLOTS <- TRUE

#-----#
```
#Settings check
Check if the values that were given in the settings are allowed
```{r}
Fun_settings_check2()
```
#Load the data from the previous file
```{r}
data <- readRDS(paste0(OUTPUT_FILE_LOCATION,"/GSE270199_imputed.rds"))
metadata <- readRDS(paste0(OUTPUT_FILE_LOCATION,"/GSE270199_redmeta.rds"))
```




```{r}
#Get the patient identifiers
healthy_group <- rownames(metadata)[grep("healthy", metadata[,"disease_state"])]
EDS_group <- rownames(metadata)[grep("EDS-HT", metadata[,"disease_state"])]


sum <- 0
temp <- NULL
for (i in 1:nrow(data)){
x <- (as.double(data[i,healthy_group]))
y <- (as.double(data[i,EDS_group]))
res <- wilcox.test(x, y, paired = FALSE, alternative = "two.sided")
if(res$p.value < 0.05){
  sum <- sum + 1
  temp <- c(temp, i)}
}
print(sum)
tempSTRING <- rownames(data[temp,])

#Toss that into Cytoscape just for curiousity


#Check that Cytoscape can be reached
cytoscapePing()

#If this section of code is re-run then the code runs into issues, to prevent this, remove if there is a networkSuid in the environment already and reload it
if (exists("networkSuid")){
  cat("networkSuid was already made, removing it and remaking it to prevent issues with the code rerunning.\n")
  RCy3::deleteNetwork(networkSuid)
  rm(networkSuid)
  cat("networkSuid has been deleted and will be remade.\n")
} else if (!exists("networkSuid")){
  cat("networkSuid does not exist in the environment, creating new network. \n")
}


#Build the command to send to Cytoscape
cmd <- paste0(
  'string protein query ',
  'query="', tempSTRING, '" ',
  'networkType=full STRING network ',
#  'networkType=physical subnetwork ',
  'cutoff=0.7 ',
  'newNetName=Base Network ',
  'species="Homo sapiens" ',
  'limit=0 ',
  'showFlatNodeDesign=true ',
  'showSTRINGstyleLabels=false ',
  'showStructureImages=false'
  )

#Send the command to Cytoscape to build the network
RCy3::commandsRun(cmd)

#Get the network's SUID
networkSuid = getNetworkSuid()



```








#Advice from Martina:
(to make it so that I can use this project for MSB1014 as well)
* Remove controls and only compare the patients
* Get the top features from a PCA to build the background network 
* Similarity network of the patients
  * Correlate the patients into 2 or 3 clusters to find patient groups
  
#Advice from Agi:
For two independent groups you should use the Mann–Whitney U test (Wilcoxon rank-sum test) can be applied to each feature. In addition to the p-value, report the fold change (FC) to describe the magnitude and direction of the difference. Since you have outliers use median instead of mean.  Because many features are tested please correct the  p-values for multiple testing using the Benjamini–Hochberg FDR procedure.

#Plan:
* Get a similarity network for the patients
* Perform Wilcoxon rank-sum test on each feature using the median
  --> p-value
  --> FC
  * Correct using FDR
  




