---
title: "MSB1015 - GSE270199 Processing"
author: "Anne"
date: "2026-10"
output: html_document
---
#Loading the data
Make sure that the previous file has been run and has saved the rds files (GSE270199_imputed.rds & GSE270199_redmeta.rds found in the OUTPUTS location)

#Clear the global environment
  Clear the environment to reduce the chance of issues on reruns
```{r clear_global}
#rm(list=ls())
```
#Setting the working directory (wd)
Other files are assumed to be located in the same, or downstream maps from the location of this file. The location of this file is used as a reference point to find the other files. 
```{r working_directory}
#download and load the package for setting the working directory to the current file
if (!require("rstudioapi", quietly = TRUE))
  install.packages("rstudioapi")
library(rstudioapi)  

#Set the wd to the location of this file  
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```
#Library installations and loading 
As I will be using a project (renv) for the MSB1015 project, I check first that the packages that need to be installed are installed.
```{r library_R_file, include=FALSE}
#For cleanliness the checks, installing and loading of the various libraries is in a separate .R file. This file is assumed to be in the same map as this file (in the wd).
source("libraries.R")
```

```{r library_installation}
#Use the function to install the libraries
Fun_install_libraries()
```

```{r library_loading}
#Use the function to load the libraries
Fun_load_libraries()
```
#Load in the checks R file
There are some functions to run checks for various things in this separate .R file.
These are not in the main code as they create quite a bit of bloat that makes the code more difficult to read.
These are not reusable functions
```{r checks}
source("checks.R")
```
#Load in the custom_functions R file
There are some functions to run checks for various things in this separate .R file.
These are not in the main code to make the main script less bloated and the custom functions easier to find (instead of all of them being spread over the Rmd file).
```{r custom_functions}
source("custom_functions.R")
```
#File locations
Set the locations of the files that are loaded in and the locations on where the outputs will go. 
This code will also do a check that the given locations are findable. 
```{r file_locations}
#Make sure to save the wd so that it is easy to use to set file locations
working_directory <- getwd()

#The default INPUT_FILE_LOCATION is in a map where this file is also saved.
INPUT_FILE_LOCATION <- paste0(working_directory,"/INPUTS")
#The default OUTPUT_FILE_LOCATION is in a map where this file is also saved.
OUTPUT_FILE_LOCATION <- paste0(working_directory,"/OUTPUTS")

#-----#

#Make sure that the file locations can be found and in the case of the outputs create a dir if it doesn't exist
Fun_in_out_check()

```
#Settings
Any global settings that are used are set here.
*Please make sure the settings are the same between the files when applicable* 
```{r settings}
#-----#
#In multiple sections:
  #During testing it was often beneficial to run the whole file. Preventing the plots from being printed allows the file to run slightly faster.
  #This setting is mostly used for testing purposes. 
#TAG: add into the settings check
RUN_PLOTS <- TRUE

#-----#
```
#Settings check
Check if the values that were given in the settings are allowed
```{r}
Fun_settings_check2()
```
#Load the data from the previous file
Now named data and metadata
```{r}
data <- readRDS(paste0(OUTPUT_FILE_LOCATION,"/GSE270199_imputed.rds"))
metadata <- readRDS(paste0(OUTPUT_FILE_LOCATION,"/GSE270199_redmeta.rds"))
```

#Build a similarity network and cut into clusters
Build a similarity network, 
  first for all the samples for exploration
  then with the '700 sample' removed as it simply acts too much like an outlier
  then with just the EDS samples, as that is what we want to cluster
  then with the clusters visualized
```{r similarity_network}
#Get some group information that can be used for color labels
healthy_group <- rownames(metadata)[grep("healthy", metadata[,"disease_state"])]
EDS_group <- rownames(metadata)[grep("EDS-HT", metadata[,"disease_state"])]
Male_group <- rownames(metadata)[grep("Male", metadata[,"inferred_sex"])]
Female_group <- rownames(metadata)[grep("Female", metadata[,"inferred_sex"])]

if (RUN_PLOTS){
  #Using all the data
  Fun_heatmap(data_matrix = t(data),
              title = "Heatmap of all the data",
              col_side = "sex",
              xlab = "Sex",
              row_side = "disease",
              ylab = "Disease")
}
if (RUN_PLOTS){
  #Without GSM8336700
    #700 keeps acting like an outlier, wanted to see how the similarity would look without it
  Fun_heatmap(data_matrix = t(data[,colnames(data) != "GSM8336700"]),
              title = "Heatmap without GSM8336700",
              col_side = "sex",
              xlab = "Sex",
              row_side = "disease",
              ylab = "Disease")
}
if (RUN_PLOTS){
  #With only the EDS samples
  Fun_heatmap(data_matrix = t(data[,EDS_group]),
              title = "Heatmap - Only EDS",
              col_side = "sex",
              xlab = "Sex",
              row_side = "disease",
              ylab = "Disease")
}
if (RUN_PLOTS){
  #With 4 clusters
  Fun_heatmap(data_matrix = t(data[,EDS_group]),
              clusters = 4,
              title = "Heatmap - Clustered EDS",
              col_side = "cluster",
              xlab = "Cluster",
              row_side = "sex",
              ylab = "Sex")
}

```

```{r clusters}
#Use the same code inside the function to get the clusters out and into the metadata
#Compute similarity using Euclidean
#Euclidean as I want simply the distance measure between the samples
sim_obj <- simil(data_matrix, method = "Euclidean")
#Convert to a matrix 
sim_matrix <- as.matrix(sim_obj)
#Set the self-correlation to NA for nicer visualization
sim_matrix[sim_matrix == 0] <- NA
#Get the dissimilarity
  #The default of dist is also Euclidean, just like in the simil 
dist_matrix <- dist(sim_matrix)
#Get the hierarchical clustering
hc <- hclust(dist_matrix)
#Get the 4 clusters
clusters_k <- cutree(hc, 4)
clusters_k <- as.data.frame(clusters_k)
#Set the controls to cluster 0
clusters_k[healthy_group,] <- 0

#TAG: CONTINUE HERE
#TAG: Go back to metadata in the previous file and remove the Row.names to prevent the duplicate issue here
temp <- merge(metadata, clusters_k, by = "row.names")
rownames(temp) <- temp$Row.names
temp$Row.names <- NULL

```




#Experimenting
Experimenting with the wilcox.test
```{r}
#Get the patient identifiers
healthy_group <- rownames(metadata)[grep("healthy", metadata[,"disease_state"])]
EDS_group <- rownames(metadata)[grep("EDS-HT", metadata[,"disease_state"])]


sum <- 0
temp <- NULL
for (i in 1:nrow(data)){
  
  
  
x <- (as.double(data[i,healthy_group]))
y <- (as.double(data[i,EDS_group]))
res <- wilcox.test(x, y, paired = FALSE, alternative = "two.sided")
if(res$p.value < 0.05){
  sum <- sum + 1
  temp <- c(temp, i)}
}
print(sum)
tempSTRING <- rownames(data[temp,])

#Toss that into Cytoscape just for curiousity


#Check that Cytoscape can be reached
cytoscapePing()

#If this section of code is re-run then the code runs into issues, to prevent this, remove if there is a networkSuid in the environment already and reload it
if (exists("networkSuid")){
  cat("networkSuid was already made, removing it and remaking it to prevent issues with the code rerunning.\n")
  RCy3::deleteNetwork(networkSuid)
  rm(networkSuid)
  cat("networkSuid has been deleted and will be remade.\n")
} else if (!exists("networkSuid")){
  cat("networkSuid does not exist in the environment, creating new network. \n")
}


#Build the command to send to Cytoscape
cmd <- paste0(
  'string protein query ',
  'query="', tempSTRING, '" ',
  'networkType=full STRING network ',
#  'networkType=physical subnetwork ',
  'cutoff=0.7 ',
  'newNetName=Base Network ',
  'species="Homo sapiens" ',
  'limit=0 ',
  'showFlatNodeDesign=true ',
  'showSTRINGstyleLabels=false ',
  'showStructureImages=false'
  )

#Send the command to Cytoscape to build the network
RCy3::commandsRun(cmd)

#Get the network's SUID
networkSuid = getNetworkSuid()



```








#Advice from Martina:
(to make it so that I can use this project for MSB1014 as well)
* Remove controls and only compare the patients
* Get the top features from a PCA to build the background network 
* Similarity network of the patients
  * Correlate the patients into 2 or 3 clusters to find patient groups
  
#Advice from Agi:
For two independent groups you should use the Mann–Whitney U test (Wilcoxon rank-sum test) can be applied to each feature. In addition to the p-value, report the fold change (FC) to describe the magnitude and direction of the difference. Since you have outliers use median instead of mean.  Because many features are tested please correct the  p-values for multiple testing using the Benjamini–Hochberg FDR procedure.

#Advice from Agi:
(data[feature, group])
median is for the FC calculation between the groups

Looking at the variability of the genes
You will not expect the variability to go to an unreasenably high value

imput data --> use PCA as dimentionallity reduction --> use the scores for similarity / other clustering / affinity propagation clustering (available in R), creates the clusters
PCA still contains the features but only contains (ex. 90%) get clusters, go back to real data and color code the cluster to what was found in the reduction on the real data
CHECK THAT THERE IS NO STRUCTURE IN THE RESIDUALS

#Plan:
* Get a similarity network for the patients
* Perform Wilcoxon rank-sum test on each feature using the median
  --> p-value
  --> FC
  * Correct using FDR
  




