---
title: "MSB1015 - GEOD77758 Loading and Cleaning"
author: "Anne Savelkoul - i6326700"
date: "2026-09"
---

#Downloading the data
Download GSE270199_COL3A1_normalizedCounts.txt.gz from https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE270199. Unzip it and place it into the INPUT_FILE_LOCATION you specify in File locations. 

Download the Metadata of GSE270199_COL3A1_normalizedCounts.txt.gz from https://www.ncbi.nlm.nih.gov/Traces/study/?acc=PRJNA1125404&o=acc_s%3Aa. Unzip it and place it into the INPUT_FILE_LOCATION you specify in File locations. 
*RENAME IT TO "GSE270199_SraRunTable.csv"* instead of "SraRunTable.csv" to prevent multiple files being names "SraRunTable.csv".

Download the 22 raw data files from https://www.ebi.ac.uk/biostudies/ArrayExpress/studies/E-GEOD-77758?query=%22hypermobility%20syndrome%22. Unzip the files and place them into the INPUT_FILE_LOCATION you specify in File locations.

Download the Metadata of GEOD77758 from https://www.ebi.ac.uk/biostudies/ArrayExpress/studies/E-GEOD-77758?query=%22hypermobility%20syndrome%22 as well. Look in "MAGE-TAB Files" and download the file named "E-GEOD-77758.sdrf.txt". 

*Unless specified otherwise, the code attempts to load in the default names of the files. If the names are changed in other ways then adjust the code in Load data.*



#Setting the working directory (wd)
Other files are assumed to be located in the same, or downstream maps from the location of this file. The location of this file is used as a reference point to find the other files. 
```{r working_directory}
#download and load the package for setting the working directory to the current file
if (!require("rstudioapi", quietly = TRUE))
  install.packages("rstudioapi")
library(rstudioapi)  

#Set the wd to the location of this file  
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```
#Library installations and loading 
As I will be using a project (renv) for the MSB1015 project, I check first that the packages that need to be installed are installed.
```{r library_R_file, include=FALSE}
#For cleanliness the checks, installing and loading of the various libraries is in a separate .R file. This file is assumed to be in the same map as this file (in the wd).
source("libraries.R")
```
```{r library_installation}
#Use the function to install the libraries
Fun_install_libraries()
```
```{r library_loading}
#Use the function to load the libraries
Fun_load_libraries()
```
#Load in the checks R file
```{r checks}
#There are some functions to run checks for various things in this separate .R file.
#These are not in the main code as they create quite a bit of bloat that makes the code more difficult to read.
#These are not reusable functions!
source("checks.R")
```
#Load in the custom_functions R file
```{r custom_functions}
#There are some functions to run checks for various things in this separate .R file.
#These are not in the main code to make the main script less bloated and the custom functions easier to find (instead of all of them being spread over the Rmd file).
source("custom_functions.R")
```
#File locations
Set the locations of the files that are loaded in and the locations on where the outputs will go. 
This code will also do a check that the given locations are findable. 
```{r file_locations}
#Make sure to save the wd so that it is easy to use to set file locations
working_directory <- getwd()

#The default INPUT_FILE_LOCATION is in a map where this file is also saved.
INPUT_FILE_LOCATION <- paste0(working_directory,"/INPUTS")
#The default OUTPUT_FILE_LOCATION is in a map where this file is also saved.
OUTPUT_FILE_LOCATION <- paste0(working_directory,"/OUTPUTS")

#-----#

#Make sure that the file locations can be found and in the case of the outputs create a dir if it doesn't exist
Fun_in_out_check()

```
#Settings
Any global settings that are used are set here. 
```{r settings}
#-----#
#In section: remove_missing
  #Missing values are removed from GSE270199, these are the thresholds that are permitted for the gene to be kept. 
#Set the percentage (0 to 100) of zeros that are allowed in the healthy group
  #There are 4 values, setting it to 25 means that 3/4 need to be non-zero
allow_zeros_in_healthy_percent <- 25
#Set the percentage (0 to 100) of zeros that are allowed in the EDS group
  #There are 10 values, setting it to 20 means that 8/10 need to be non-zero
allow_zeros_in_EDS_percent <- 20
#-----#
```
#Settings check
Check if the values that were given in the settings are allowed
```{r}
Fun_settings_check()
```
#Load GEOD77758 data
The data is loaded here. 
```{r load_data_GEOD77758}

#Get all the files with a .CEL extention and HuGenes in the name
cels.HuGene <- list.files(paste0(INPUT_FILE_LOCATION,"/"), pattern = c("HuGene", "CEL"))
#Get all the files with a .CEL extention and mi in the name
cels.miRNA <- list.files(paste0(INPUT_FILE_LOCATION,"/"), pattern = c("miRNA", "CEL"))

#Check if the files could be found
Fun_GEOD77758_findability_check()


#-----# 

#Read the .CELL files
GEOD77758.HuGene.all <- read.celfiles(verbose=FALSE, filenames=paste0(INPUT_FILE_LOCATION,"/", cels.HuGene))

#Use rma to get the assayData expression out of the GEOD77758.all
GEOD77758.HuGene.rma <- rma(GEOD77758.HuGene.all, background=TRUE, normalize=TRUE, subset=NULL, target="core")

GEOD77758.HuGene <- GEOD77758.HuGene.rma@assayData$exprs

#Simplify the colnames
GEOD77758.colnames <- colnames(GEOD77758.HuGene)

GEOD77758.colnames <- gsub(pattern = "_EDS-HT_Patient_.*_HuGene-1_0-st-v1_.CEL", replacement = "", x = GEOD77758.colnames)
GEOD77758.colnames <- gsub(pattern = "_Control_.*_HuGene-1_0-st-v1_.CEL", replacement = "", x = GEOD77758.colnames)

colnames(GEOD77758.HuGene) <- GEOD77758.colnames

#Remove cels.HuGene as it is no longer used
rm(cels.HuGene)
#Remove GEOD77758.HuGene.all and GEOD77758.HuGene.rma as they are no longer used
rm(GEOD77758.HuGene.all, GEOD77758.HuGene.rma)
#Remove GEOD77758.colnames as it is no longer used
rm(GEOD77758.colnames)

#-----# 

#Read the .CELL files
GEOD77758.miRNA.all <- read.celfiles(verbose=FALSE, filenames=paste0(INPUT_FILE_LOCATION,"/", cels.miRNA))

#Use rma to get the assayData expression out of the GEOD77758.all
GEOD77758.miRNA.rma <- rma(GEOD77758.miRNA.all, background=TRUE, normalize=TRUE, subset=NULL)

GEOD77758.miRNA <- GEOD77758.miRNA.rma@assayData$exprs

#Simplify the colnames
GEOD77758.colnames <- colnames(GEOD77758.miRNA)

GEOD77758.colnames <- gsub(pattern = "_EDS-HT_Patient_.*_miRNA-3_0_.CEL", replacement = "", x = GEOD77758.colnames)
GEOD77758.colnames <- gsub(pattern = "_Control_.*_miRNA-3_0_.CEL", replacement = "", x = GEOD77758.colnames)

colnames(GEOD77758.miRNA) <- GEOD77758.colnames

#Remove cels.miRNA as it is no longer used
rm(cels.miRNA)
#Remove GEOD77758.miRNA.all and GEOD77758.miRNA.rma as they are no longer used
rm(GEOD77758.miRNA.all, GEOD77758.miRNA.rma)
#Remove GEOD77758.colnames as it is no longer used
rm(GEOD77758.colnames)

#-----# 

#Load the metadata for GEOD77758
GEOD77758.meta <- read.delim(paste0(INPUT_FILE_LOCATION,"/E-GEOD-77758.sdrf.txt"))
rownames(GEOD77758.meta) <- GEOD77758.meta[,"Assay.Name"]

#-----# 

#Split the GEOD77758.meta into GEOD77758.HuGene.meta and GEOD77758.miRNA.meta
GEOD77758.HuGene.meta <- GEOD77758.meta[grep("HuGene", GEOD77758.meta[,"Array.Data.File"]),]
GEOD77758.miRNA.meta <- GEOD77758.meta[grep("miRNA", GEOD77758.meta[,"Array.Data.File"]),]

#Check that the people in GEOD77758 are also in GEOD77758.meta and that the people in GEOD77758 are also in GEOD77758.meta
Fun_GEOD77758_meta_check()

```

#Metadata
Check what is available in the metadata and remove un-relevant information / add information that could be useful.
```{r metadata}
#Check metadata

#Break up the output
cat("\n-----\n")
#-----# 

#Create reduced metadata (redmeta) that removes columns that only contain a single thing
keep <- NULL
for (column in 1:ncol(GEOD77758.HuGene.meta)) {
  keep <- c(keep,length(unique(GEOD77758.HuGene.meta[,column])) != 1)
}
GEOD77758.HuGene.redmeta <- GEOD77758.HuGene.meta[,keep]

#Remove ones that are fully unique (cannot check for any batch effects on that anyway)
keep <- NULL
for (column in 1:ncol(GEOD77758.HuGene.redmeta)) {
  keep <- c(keep,length(unique(GEOD77758.HuGene.redmeta[,column])) != nrow(GEOD77758.HuGene.redmeta))
}

GEOD77758.HuGene.redmeta <- GEOD77758.HuGene.redmeta[,keep]

#Manually remove colums that are unrelevant
cat("Remaining columns after automatic removal in GEOD77758.HuGene.redmeta:\n", colnames(GEOD77758.HuGene.redmeta),"\n")
cat("\"Characteristics..disease.state.\" and \"FactorValue..disease.state.\" contain the same data, only keep \"Characteristics..disease.state.\"\n")
GEOD77758.HuGene.redmeta <- GEOD77758.HuGene.redmeta["Characteristics..disease.state."]

#Break up the output
cat("\n-----\n")

#Rename the remaining colnames for ease of use
colnames(GEOD77758.HuGene.redmeta) <- "disease_state"
cat("GEOD77758.HuGene.redmeta's \"Characteristics..disease.state.\" has been renamed to \"disease_state\" to align with the other metadata files.\n")

#Break up the output
cat("\n-----\n")
#-----# 

#Create reduced metadata (redmeta) that removes columns that only contain a single thing
keep <- NULL
for (column in 1:ncol(GEOD77758.miRNA.meta)) {
  keep <- c(keep,length(unique(GEOD77758.miRNA.meta[,column])) != 1)
}
GEOD77758.miRNA.redmeta <- GEOD77758.miRNA.meta[,keep]

#Remove ones that are fully unique (cannot check for any batch effects on that anyway)
keep <- NULL
for (column in 1:ncol(GEOD77758.miRNA.redmeta)) {
  keep <- c(keep,length(unique(GEOD77758.miRNA.redmeta[,column])) != nrow(GEOD77758.miRNA.redmeta))
}

GEOD77758.miRNA.redmeta <- GEOD77758.miRNA.redmeta[,keep]

#Manually remove colums that are unrelevant
cat("Remaining columns after automatic removal in GEOD77758.miRNA.redmeta:\n", colnames(GEOD77758.miRNA.redmeta),"\n")
cat("\"Characteristics..disease.state.\" and \"FactorValue..disease.state.\" contain the same data, only keep \"Characteristics..disease.state.\"\n")
GEOD77758.miRNA.redmeta <- GEOD77758.miRNA.redmeta["Characteristics..disease.state."]

#Break up the output
cat("\n-----\n")

#Rename the remaining colnames for ease of use
colnames(GEOD77758.miRNA.redmeta) <- "disease_state"
cat("GEOD77758.miRNA.redmeta's \"Characteristics..disease.state.\" has been renamed to \"disease_state\" to align with the other metadata files.\n")

#Break up the output
cat("\n-----\n")
#-----#

#Remove keep and rename as they are no longer needed
rm(keep, rename)

```

#Missing values
Check if the loaded data has any missing values. 
```{r check_missing_1}
#Break up the output
cat("\n-----\n")

if (!(NA %in% GEOD77758.HuGene)){
  cat("There are no \"NA\" values in GEOD77758.HuGene. \n")
} else if (NA %in% GEOD77758.HuGene){
  warning("There are \"NA\" values in GEOD77758.HuGene. \n")
} else {
  warning("Something unknown went wrong\n")
}

#Check how many zeros are in the data
cat(paste0("There are ", sum(GEOD77758.HuGene == 0), " zero values in GEOD77758.HuGene \n",
           "That is ", 
           round((sum(GEOD77758.HuGene == 0)/(ncol(GEOD77758.HuGene) * nrow(GEOD77758.HuGene)))*100, 2), 
           "% of the data.\n"))

#Check what the smallest value is
cat(paste0("The smallest value in GEOD77758.HuGene (ignorming zeros) is ", min(GEOD77758.HuGene[GEOD77758.HuGene != 0]),".\n"))

#Break up the output
cat("\n-----\n")

if (!(NA %in% GEOD77758.miRNA)){
  cat("There are no \"NA\" values in GEOD77758.miRNA. \n")
} else if (NA %in% GEOD77758.miRNA){
  warning("There are \"NA\" values in GEOD77758.miRNA. \n")
} else {
  warning("Something unknown went wrong\n")
}

#Check how many zeros are in the data
cat(paste0("There are ", sum(GEOD77758.miRNA == 0), " zero values in GEOD77758.miRNA \n",
           "That is ", 
           round((sum(GEOD77758.miRNA == 0)/(ncol(GEOD77758.miRNA) * nrow(GEOD77758.miRNA)))*100, 2), 
           "% of the data.\n"))

#Check what the smallest value is
cat(paste0("The smallest value in GEOD77758.miRNA (ignorming zeros) is ", min(GEOD77758.miRNA[GEOD77758.miRNA != 0]),".\n"))

#Break up the output
cat("\n-----\n")

if (!(NA %in% GEOD77758.meta)){
  cat("There are no \"NA\" values in GEOD77758.meta. \n")
} else if (NA %in% GEOD77758.meta){
  warning("There are \"NA\" values in GEOD77758.meta. \n")
} else {
  warning("Something unknown went wrong\n")
}

#Break up the output
cat("\n-----\n")

```





