---
title: "MSB1015 - GSE270199 Loading and Cleaning"
author: "Anne Savelkoul - i6326700"
date: "2026-09"
output: html_document

---

#Downloading the data
Download GSE270199_COL3A1_normalizedCounts.txt.gz from https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE270199. Unzip it and place it into the INPUT_FILE_LOCATION you specify in File locations. 

Download the Metadata of GSE270199_COL3A1_normalizedCounts.txt.gz from https://www.ncbi.nlm.nih.gov/Traces/study/?acc=PRJNA1125404&o=acc_s%3Aa. Unzip it and place it into the INPUT_FILE_LOCATION you specify in File locations. 
*RENAME IT TO "GSE270199_SraRunTable.csv"* instead of "SraRunTable.csv" to prevent multiple files being names "SraRunTable.csv".

*Unless specified otherwise, the code attempts to load in the default names of the files. If the names are changed in other ways then adjust the code in Load data.*

#Note:
If there are sections with a "#TAG: [description]" then that means that that code has been tagged for needing changes made / to make things that might need to be changed easier to find. 

#Clear the global environment
  Clear the environment to reduce the chance of issues on reruns
```{r clear_global}
rm(list=ls())
```

#Setting the working directory (wd)
Other files are assumed to be located in the same, or downstream maps from the location of this file. The location of this file is used as a reference point to find the other files. 
```{r working_directory}
#download and load the package for setting the working directory to the current file
if (!require("rstudioapi", quietly = TRUE))
  install.packages("rstudioapi")
library(rstudioapi)  

#Set the wd to the location of this file  
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```
#Library installations and loading 
As I will be using a project (renv) for the MSB1015 project, I check first that the packages that need to be installed are installed.
```{r library_R_file, include=FALSE}
#For cleanliness the checks, installing and loading of the various libraries is in a separate .R file. This file is assumed to be in the same map as this file (in the wd).
source("libraries.R")
```
Library installation
```{r library_installation}
#Use the function to install the libraries if they are missing.
Fun_install_libraries()
```
Library loading
```{r library_loading}
#Use the function to load the libraries
Fun_load_libraries()
```
#TAG: Remove section if I cannot get Agi's functions to run
#Load Agi's URF functions
  Adds the functions:
    script_URF
    unsupervisedRF_proximityadd
    doublecentering
```{r}
#The R session aborted very quickly after trying to play with the functions, so I am leaving Agi's functions alone for now.
#source("URF.R")
```

#Load in the checks R file
There are some functions to run checks for various things in this separate .R file.
These are not in the main code as they create quite a bit of bloat that makes the code more difficult to read.
These are not reusable functions!
```{r checks}
source("checks.R")
```
#Load in the custom_functions R file
There are some functions to run checks for various things in this separate .R file.
These are not in the main code to make the main script less bloated and the custom functions easier to find (instead of all of them being spread over the Rmd file).
```{r custom_functions}
source("custom_functions.R")
```
#File locations
Set the locations of the files that are loaded in and the locations on where the outputs will go. 
This code will also do a check that the given locations are findable. 
```{r file_locations}
#Make sure to save the wd so that it is easy to use to set file locations
working_directory <- getwd()

#The default INPUT_FILE_LOCATION is in a map where this file is also saved.
INPUT_FILE_LOCATION <- paste0(working_directory,"/INPUTS")
#The default OUTPUT_FILE_LOCATION is in a map where this file is also saved.
OUTPUT_FILE_LOCATION <- paste0(working_directory,"/OUTPUTS")

#-----#

#Make sure that the file locations can be found and in the case of the outputs create a dir if it doesn't exist
Fun_in_out_check()

```
#Settings
Any global settings that are used are set here. 
```{r settings}
#-----#
#In multiple sections:
  #During testing it was often beneficial to run the whole file. Preventing the plots from being printed allows the file to run slightly faster.
  #This setting is mostly used for testing purposes. 
RUN_PLOTS <- TRUE

#-----#
#In section: inferred_sex
#TAG: Add description
#TAG: Add to check
QUERY_ENSEMBL <- FALSE
#TAG: Add description
#TAG: Add to check
OVERRIDE_Y_CHRO_SAVED_FILE <- TRUE
  
#-----#
#In section: remove_missing
  #Missing values are removed from GSE270199, these are the thresholds that are permitted for the gene to be kept. 
#Set the percentage (0 to 100) of zeros that are allowed in the healthy group
  #There are 4 values, setting it to 25 means that 3/4 need to be non-zero
allow_zeros_in_healthy_percent <- 25
#Set the percentage (0 to 100) of zeros that are allowed in the EDS group
  #There are 10 values, setting it to 20 means that 8/10 need to be non-zero
allow_zeros_in_EDS_percent <- 20
#Set which filtering you would like to continue with.
  #Options:
    #"either" --> Either the EDS or the healthy group needs to pass the thresholds for the gene to be kept.
    #"both" --> Both the EDS and the healthy group need to pass the thresholds for the gene to be kept.
    #"difference" --> The difference between the previous two options.
FILTERSETTING <- "either"

#-----#
#In section: imputation
#Set if you would like to impute the data (TRUE/FALSE) to 80% of the mean of non-zero values in the group.
IMPUTE_MISSING <- TRUE
#Set if you would like to impute the data (TRUE/FALSE) to zero if enough data within that group is missing. 
  #How much data needs to be missing is set in the zeros_for_imputation settings below. 
IMPUTE_ZEROS <- TRUE
#When you impute, how many samples withing the group need to be missing for the remainder to be set to zero.
  #Ex: zeros_for_imputation_EDS at 8 means that 8 of the 10 patient samples
  # need to be zero for the remaining two to also be set to zero.
  #Ex: zeros_for_imputation_healthy at 3 means that 3 of the 4 healthy samples
  # need to be zero for the remaining one to also be set to zero.
#TAG: turn this into percentages if time allows
zeros_for_imputation_EDS <- 8
zeros_for_imputation_healthy <- 3


```
#Settings check
Check if the values that were given in the settings are allowed
```{r}
Fun_settings_check1()
```
#Load GSE270199 data
The data is loaded here. 
```{r load_data_GSE270199}
#Check that the required inputs can be found, otherwise stop the code from running
Fun_GSE270199_findability_check()

#-----# 

#Load GSE270199 and fix up the rownames
GSE270199 <- read.delim(paste0(INPUT_FILE_LOCATION,"/GSE270199_COL3A1_normalizedCounts.txt"))
rownames(GSE270199) <- GSE270199[,1]
GSE270199 <- GSE270199[,-1]

#Print a message if GSE270199 exists in the environment (meaning that it was loaded)
if(exists("GSE270199")) cat("GSE270199 has been loaded \n")

#-----# 

#Load the metadata of GSE270199 and fix up the rownames
#I do not remove the Library.Name from the metadata file in case I want to use it for something later
GSE270199.meta <- read.csv(paste0(INPUT_FILE_LOCATION,"/GSE270199_SraRunTable.csv"))
rownames(GSE270199.meta) <- GSE270199.meta[,"Library.Name"]

#Print a message if GSE270199 exists in the environment (meaning that it was loaded)
if(exists("GSE270199.meta")) cat("GSE270199.meta has been loaded \n")

#-----# 

#By default the names of the patients in the normalizedCounts and the metadata do not line up. 
# This is not in the metadata file and needs to be manually adjusted 
# Manually copied the aligning names from https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE270199 and rename the colnames one by one before applying them.

#Get the origninal colnames
GSE270199.colnames <- colnames(GSE270199)

#Replace them with the ones that line up with the metadata
GSE270199.colnames <- gsub(pattern = "C3B_healthy", replacement = "GSM8336700", x = GSE270199.colnames)
GSE270199.colnames <- gsub(pattern = "C2B_healthy", replacement = "GSM8336701", x = GSE270199.colnames)
GSE270199.colnames <- gsub(pattern = "RR_healthy", replacement = "GSM8336702", x = GSE270199.colnames)
GSE270199.colnames <- gsub(pattern = "C1A_healthy", replacement = "GSM8336703", x = GSE270199.colnames)
GSE270199.colnames <- gsub(pattern = "X20.3156_pat", replacement = "GSM8336704", x = GSE270199.colnames)
GSE270199.colnames <- gsub(pattern = "X20.3154_pat", replacement = "GSM8336705", x = GSE270199.colnames)
GSE270199.colnames <- gsub(pattern = "X20.4456_pat", replacement = "GSM8336706", x = GSE270199.colnames)
GSE270199.colnames <- gsub(pattern = "VL_pat", replacement = "GSM8336707", x = GSE270199.colnames)
GSE270199.colnames <- gsub(pattern = "GAU_pat", replacement = "GSM8336708", x = GSE270199.colnames)
GSE270199.colnames <- gsub(pattern = "X19.1597_pat", replacement = "GSM8336709", x = GSE270199.colnames)
GSE270199.colnames <- gsub(pattern = "SF_pat", replacement = "GSM8336710", x = GSE270199.colnames)
GSE270199.colnames <- gsub(pattern = "X22.0547_pat", replacement = "GSM8336711", x = GSE270199.colnames)
GSE270199.colnames <- gsub(pattern = "X22.0548_pat", replacement = "GSM8336712", x = GSE270199.colnames)
GSE270199.colnames <- gsub(pattern = "X20.4222_pat", replacement = "GSM8336713", x = GSE270199.colnames)
	
#Apply them to the colnames
colnames(GSE270199) <- GSE270199.colnames

#Remove GSE270199.colnames from the environment as it is no longer used
rm(GSE270199.colnames)
	
#Check that the people in GSE270199 are also in GSE270199.meta
#Modify this check if time allows so it can be reused for the other data if it gets used.
Fun_GSE270199_meta_check()
```
#Metadata
Check what is available in the metadata and remove un-relevant information / add information that could be useful.
```{r metadata}
#Check metadata

#Break up the output
cat("\n-----\n")
#Create reduced metadata (redmeta) that removes columns that only contain a single thing
keep <- NULL
for (column in 1:ncol(GSE270199.meta)) {
  keep <- c(keep,length(unique(GSE270199.meta[,column])) != 1)
}
GSE270199.redmeta <- GSE270199.meta[,keep]

#Remove ones that are fully unique (cannot check for any batch effects on that anyway)
keep <- NULL
for (column in 1:ncol(GSE270199.redmeta)) {
  keep <- c(keep,length(unique(GSE270199.redmeta[,column])) != nrow(GSE270199.redmeta))
}

GSE270199.redmeta <- GSE270199.redmeta[,keep]

#Manually remove colums that are unrelevant
cat("Remaining columns after automatic removal in GSE270199.redmeta:\n", colnames(GSE270199.redmeta),"\n")
cat("The only one considered interesting is \"disease_state\".\n")
GSE270199.redmeta <- GSE270199.redmeta["disease_state"]

#Break up the output
cat("\n-----\n")

#Rename the values inside the disease_state for ease of use
  #EDS patient --> EDS-HT/JHS
  #healthy control --> healthy
rename <- sub("EDS patient","EDS-HT/JHS", GSE270199.redmeta[,"disease_state"])	
rename <- sub("healthy control","healthy", rename)
GSE270199.redmeta[,"disease_state"] <- rename
#TAG: If the other dataset ever ends up being used anyway, change the feedback below here
cat("The values inside GSE270199.redmeta[\"disease_state\"] have been renamed to align with the data in the metadata files of the other dataset (This other dataset remained unused):\n\tEDS patient --> EDS-HT/JHS\n\thealthy control --> healthy\n")

#Break up the output
cat("\n-----\n")
#-----# 

#Remove keep and rename as they are no longer needed
rm(keep, rename, column)

```
#Detect Y chromosome markers to get male/female information for the metadata
```{r inferred_sex}
#Only query the ensemble if the file to be loaded in doesn't exist or if the QUERY_ENSEMBL setting is set to TRUE

if (QUERY_ENSEMBL | !(file.exists(paste0(OUTPUT_FILE_LOCATION,"/genes_y_chromosome.rds")))){
  #Set-up ensembl API connection
  ensembl_db <- useEnsembl(biomart = "genes")
  ensembl_dataset <- useDataset(dataset = "hsapiens_gene_ensembl",
                                mart = ensembl_db)
  
  #Get genes in chromosome Y from ensembl
  genes_y_chromosome <- getBM(
    attributes = c("ensembl_gene_id"),
    filters = "chromosome_name",
    values = "Y",
    mart = ensembl_dataset
  )
  
  #Save the genes_y_chromosome.rds if the genes_y_chromosome.rds is not found in the OUTPUTS folder or if the OVERRIDE_Y_CHRO_SAVED_FILE setting to override the current file is set to TRUE
  if (!(file.exists(paste0(OUTPUT_FILE_LOCATION,"/genes_y_chromosome.rds"))) | OVERRIDE_Y_CHRO_SAVED_FILE){
  #The ensemble database was being temperamental, so save the results to the output so that those can be loaded instead if needed
  saveRDS(genes_y_chromosome, paste0(OUTPUT_FILE_LOCATION,"/genes_y_chromosome.rds"))
  }
} else {
  genes_y_chromosome <- readRDS(paste0(OUTPUT_FILE_LOCATION,"/genes_y_chromosome.rds"))
}

#Keep only genes that are found in the dataset
genes_y_chromosome_in_data <- genes_y_chromosome |>
  filter(ensembl_gene_id %in% rownames(GSE270199))

#Get some feedback
cat(paste0(nrow(genes_y_chromosome_in_data)," out of ", nrow(genes_y_chromosome), " of the Y chromosome genes were found in the data."))

#Get the y_genes in a way that can be used to subset the data
y_genes <- genes_y_chromosome_in_data[,1]

#As the data is pre-normalized simply taking the sum and seeing which are above zero should be enough
  #Feel free to run: 
  # colSums(GSE270199[y_genes, ])
  # to verify this
inferred_sex <- ifelse(colSums(GSE270199[y_genes, ]) == 0,
                       "Female",
                       "Male")
inferred_sex <- as.data.frame(inferred_sex)

#Add the inferred_sex to the reduced metadata
GSE270199.redmeta <- merge(GSE270199.redmeta, inferred_sex, by='row.names', all=TRUE)
rownames(GSE270199.redmeta) <- GSE270199.redmeta$Row.names

#Also get a combined disease and sex for the reduced metadata 
disease_sex <- ifelse(GSE270199.redmeta$disease_state == "EDS-HT/JHS",
                      ifelse(GSE270199.redmeta$inferred_sex == "Male",
                             "EDS_Male",
                             "EDS_Female"),
                      ifelse(GSE270199.redmeta$inferred_sex == "Male",
                             "healthy_Male",
                             "healthy_Female"))

#Add the disease_sex to the reduced metadata
GSE270199.redmeta["disease_sex"] <- disease_sex

```


#TAG: REMOVE THIS, THIS WAS A TEST ON HOW THE DATA RESPONDED TO GSM8336700 BEING REMOVED
  The answer was confusing, not in a way I want to deal with
```{r temp_test}
#GSE270199 <- GSE270199[colnames(GSE270199) != "GSM8336700"]
#
#temp <- as.data.frame(GSE270199.redmeta[grep("GSM8336700", rownames(GSE270199.redmeta), invert = TRUE),"disease_state"])
#colnames(temp) <- "disease_state"
#rownames(temp) <- rownames(GSE270199.redmeta)[grep("GSM8336700", rownames(GSE270199.redmeta), invert = TRUE)]
#
#GSE270199.redmeta <- temp
```
#Missing values
Check if the loaded data has any missing values. 
```{r check_missing_1}
#Break up the output
cat("\n-----\n")

#Check if any values are NA 
Fun_NA_check(GSE270199, "GSE270199")

#Check how many zeros are in the data
Fun_zero_check(GSE270199, "GSE270199")

#Check what the smallest value is
Fun_smallest_value(GSE270199, "GSE270199")


#Break up the output
cat("\n-----\n")

Fun_NA_check(GSE270199.meta, "GSE270199.meta")

#Break up the output
cat("\n-----\n")
```
#GSE270199 missing data processing
GSE270199 has many zeros, so those values need to be processed.
```{r remove_missing_GSE270199}
#Get the patient identifiers so that the zero removal can be done while keeping the groups in mind.
healthy_group <- rownames(GSE270199.redmeta)[grep("healthy", GSE270199.redmeta[,"disease_state"])]
EDS_group <- rownames(GSE270199.redmeta)[grep("EDS-HT", GSE270199.redmeta[,"disease_state"])]

#Convert the percentage set in the settings to a number that is used for the comparison
allow_zeros_healthy <- (length(healthy_group)*(allow_zeros_in_healthy_percent/100))

#Convert the percentage to a number that is used for the comparison
allow_zeros_EDS <- (length(EDS_group)*(allow_zeros_in_EDS_percent/100))

#Loop over GSE270199 to find the ones that have too many zeros
keep_either <- NULL
keep_both <- NULL
for (i in 1:nrow(GSE270199)){
  keep_EDS <- (sum(GSE270199[i,EDS_group] == 0) <= allow_zeros_EDS)
  keep_healthy <- (sum(GSE270199[i,healthy_group] == 0) <= allow_zeros_healthy)
  
  #It can be decided to keep 
    #Either: Either one or the other group needs to pass their threshold.
    #Both: Both of the groups need to pass their threshold.
  keep_either <- c(keep_either, keep_EDS | keep_healthy)
  keep_both <- c(keep_both, keep_EDS & keep_healthy)
}

#Get the filtered dataframes based on the keep's that were made
#Filtered of the dataframe if either of the groups pass the threshold
GSE270199.filtered_either <- GSE270199[keep_either,]
#Filtered of the dataframe if both of the groups pass the threshold
GSE270199.filtered_both <- GSE270199[keep_both,]
#Filtered of the dataframe if only one of the groups pass the threshold
GSE270199.filtered_difference <- GSE270199[(!keep_both & keep_either),]


#Remove the temp objects from the environment
rm(keep_either, keep_both, keep_EDS, keep_healthy, i)

```
#Select filtering using missingness plots
Plot all of the different data created in the script so far to compare the different types of data. 
*The missingness plots will create a warning, I elected to ignore this warning for now.* 
```{r missingness_plots}
#Create color_groups to color the labels of the samples so it is easier to see what the controls are vs the patients
color_groups <- ifelse(GSE270199.redmeta[colnames(GSE270199),"disease_state"] == "EDS-HT/JHS", "red", "black")

#Plot the missingness plots
if (RUN_PLOTS){
  Fun_missingness(GSE270199, color_groups, "GSE270199")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_missingness(GSE270199.filtered_either, color_groups, "GSE270199 filtered, minimum one passes")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_missingness(GSE270199.filtered_both, color_groups, "GSE270199 filtered, both pass")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_missingness(GSE270199.filtered_difference, color_groups, "GSE270199 difference between min one and both")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}

```
#Compare the pca plots of the different filtered data
Create the PCA objects
```{r pca_plots_1_1}
#Calculate the PCA for the non-filtered data
PCA.original <- pca(object = t(as.matrix(GSE270199)), 
                   method = "svd", nPcs = 9, center = TRUE)
#Calculate the PCA for filtered either
PCA.filtered_either <- pca(object = t(as.matrix(GSE270199.filtered_either)), 
                   method = "svd", nPcs = 9, center = TRUE)
#Calculate the PCA for filtered both
PCA.filtered_both <- pca(object = t(as.matrix(GSE270199.filtered_both)), 
                   method = "svd", nPcs = 9, center = TRUE)
#Calculate the PCA for the difference
PCA.filtered_difference <- pca(object = t(as.matrix(GSE270199.filtered_difference)), 
                   method = "svd", nPcs = 9, center = TRUE)
```
Visualize the scores of the first two PC's
```{r pca_plots_1_2}
if (RUN_PLOTS){
  Fun_plot_PCA_scores(PCA = PCA.original, 
                      metadata = GSE270199.redmeta, 
                      firstPC = "PC1", 
                      secondPC = "PC2", 
                      title = "Original PCA scores")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_plot_PCA_scores(PCA = PCA.filtered_either, 
                      metadata = GSE270199.redmeta, 
                      firstPC = "PC1", 
                      secondPC = "PC2", 
                      title = "PCA scores, either filtered")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_plot_PCA_scores(PCA = PCA.filtered_both, 
                      metadata = GSE270199.redmeta, 
                      firstPC = "PC1", 
                      secondPC = "PC2", 
                      title = "PCA scores, both filtered")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_plot_PCA_scores(PCA = PCA.filtered_difference, 
                      metadata = GSE270199.redmeta, 
                      firstPC = "PC1", 
                      secondPC = "PC2", 
                      title = "PCA scores difference between either and both filters")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
```
Visualize the loadings of the first two PC's
```{r pca_plots_1_3}
if (RUN_PLOTS){
  Fun_plot_PCA_loadings(PCA = PCA.original, 
                        firstPC = "PC1", 
                        secondPC = "PC2", 
                        title = "Original PCA scores")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_plot_PCA_loadings(PCA = PCA.filtered_either, 
                        firstPC = "PC1", 
                        secondPC = "PC2", 
                        title = "PCA scores, either filtered")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_plot_PCA_loadings(PCA = PCA.filtered_both, 
                        firstPC = "PC1", 
                        secondPC = "PC2", 
                        title = "PCA scores, both filtered")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_plot_PCA_loadings(PCA = PCA.filtered_difference, 
                        firstPC = "PC1", 
                        secondPC = "PC2", 
                        title = "PCA scores difference between either and both filters")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
```
#Biplots
Check the top most influential genes using a biplot
```{r biplots}
if (RUN_PLOTS){
  Fun_plot_biplot(PCA = PCA.original, 
                  top = 50, 
                  firstPC = "PC1", 
                  secondPC = "PC2", 
                  title = "Biplot using the top 50 genes for the non-processed data.")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_plot_biplot(PCA = PCA.filtered_either, 
                  top = 50, 
                  firstPC = "PC1", 
                  secondPC = "PC2", 
                  title = "Biplot using the top 50 genes for the either filtering.")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_plot_biplot(PCA = PCA.filtered_both, 
                  top = 50, 
                  firstPC = "PC1", 
                  secondPC = "PC2", 
                  title = "Biplot using the top 50 genes for the both filtering.")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_plot_biplot(PCA = PCA.filtered_difference, 
                  top = 50, 
                  firstPC = "PC1", 
                  secondPC = "PC2", 
                  title = "Biplot using the top 50 genes for the difference filtering.")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}

```
#Select what filtered data to continue forward with
```{r select_filtered}
#Depending on the setting specified, continue on with one of the filtered sets
if (FILTERSETTING == "either"){
  GSE270199.filtered <- GSE270199.filtered_either
  cat("Using genes that pass the threshold in either of the groups.\n")
} else if (FILTERSETTING == "both"){
  GSE270199.filtered <- GSE270199.filtered_both
  cat("Using genes that pass the threshold in both groups.\n")
} else if (FILTERSETTING == "difference"){
  GSE270199.filtered <- GSE270199.filtered_difference
  cat("Using the difference.\n")
} else {
  warning("Something unknown went wrong, stopping the code.\n")
  stop()
}
```
#Data imputation
Impute the remaining zeros within their respective groups depending on the setting provided.
```{r imputation}
if (IMPUTE_MISSING & !IMPUTE_ZEROS){
  #IMPUTE_MISSING --> Impute the missing values where possible
  #!IMPUTE_ZEROS --> Do not impute to zeros
  GSE270199.imputed <- Fun_impute_data(GSE270199.filtered, 
               allow_zeros_EDS = allow_zeros_EDS, 
               allow_zeros_healthy = allow_zeros_healthy,
               IMPUTE_TO_ZERO = FALSE
               )
  cat("Imputed missing where possible.\n")
  
} else if (IMPUTE_MISSING & IMPUTE_ZEROS){
  #IMPUTE_MISSING --> Impute the missing values where possible
  #IMPUTE_ZEROS --> Also impute to zero if the thresholds for zeros_for_imputation_EDS / zeros_for_imputation_healthy specified in settings is passed
  GSE270199.imputed <- Fun_impute_data(GSE270199.filtered, 
               allow_zeros_EDS = allow_zeros_EDS, 
               allow_zeros_healthy = allow_zeros_healthy,
               IMPUTE_TO_ZERO = TRUE,
               zeros_for_imputation_EDS = zeros_for_imputation_EDS,
               zeros_for_imputation_healthy = zeros_for_imputation_healthy
               )
  cat("Imputed missing and to zeros where possible.\n")
  
} else if (!IMPUTE_MISSING){
  #!IMPUTE_MISSING --> Do not impute anything
    #Still set it to the imputed variable name so that the following code can run it
  GSE270199.imputed <- GSE270199.filtered
  cat("Did not impute anything.\n")
  
} else {
  warning("Something unknown went wrong, stopping the code.\n")
  stop()
}
```
#Compare before and after imputation
Compare what the imputation has done to the missingness of the data.
*The implementation of the color_groups causes a warning when producing the missingness plots, this warning is ignored for now.*
```{r imputation_result}
#DUPLICATE: color_groups is created above as well, but repeated for clarity
#Create color_groups to color the labels of the samples so it is easier to see what the controls are vs the patients
color_groups <- ifelse(GSE270199.redmeta[colnames(GSE270199),"disease_state"] == "EDS-HT/JHS", "red", "black")

#Visualize the missingness plots
if (RUN_PLOTS){
  Fun_missingness(GSE270199.filtered, color_groups, "Before imputation")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_missingness(GSE270199.imputed, color_groups, "After imputation")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
```
#Missing values after correction
Check the progress of the missing value removal. 
```{r check_missing_2}
#Check how many zeros are in the data
Fun_zero_check(GSE270199.filtered, "GSE270199.filtered (before imputation)")

#Check how many zeros are in the data
Fun_zero_check(GSE270199.imputed, "GSE270199.imputed (after imputation)")
```
#Compare the pca's before and after imputation
Calculate the PCA's
```{r pca_plots_2_1}
#Calculate the PCA for the chosen filtered data
PCA.filtered <- pca(object = t(as.matrix(GSE270199.filtered)), 
                   method = "svd", nPcs = 9, center = TRUE)
#Calculate the PCA for the chosen imputed data
PCA.imputed <- pca(object = t(as.matrix(GSE270199.imputed)), 
                   method = "svd", nPcs = 9, center = TRUE)
```
Visualize the scores of the first two PCs
```{r pca_plots_2_2}
if (RUN_PLOTS){
  Fun_plot_PCA_scores(PCA = PCA.filtered, 
                      metadata = GSE270199.redmeta, 
                      firstPC = "PC1", 
                      secondPC = "PC2", 
                      title = "PCA scores before imputation")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_plot_PCA_scores(PCA = PCA.imputed, 
                      metadata = GSE270199.redmeta, 
                      firstPC = "PC1", 
                      secondPC = "PC2", 
                      title = "PCA scores after imputation")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
```
Visualize the loadings of the first two PCs
```{r pca_plots_2_3}
if (RUN_PLOTS){
  Fun_plot_PCA_loadings(PCA = PCA.filtered, 
                        firstPC = "PC1", 
                        secondPC = "PC2", 
                        title = "PCA loadings before imputation")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_plot_PCA_loadings(PCA = PCA.imputed, 
                        firstPC = "PC1", 
                        secondPC = "PC2", 
                        title = "PCA loadings after imputation")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
```
#Biplots
Compare how the biplot has changed due to the imputation
```{r biplots}
if (RUN_PLOTS){
  Fun_plot_biplot(PCA = PCA.filtered, 
                  top = 50, 
                  firstPC = "PC1", 
                  secondPC = "PC2", 
                  title = "Biplot using the top 50 genes for the filtered data.")
} else {
  cat("Plots not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_plot_biplot(PCA = PCA.imputed, 
                  top = 50, 
                  firstPC = "PC1", 
                  secondPC = "PC2", 
                  title = "Biplot using the top 50 genes for the imputed data.")
} else {
  cat("Plots not visualized due to RUN_PLOTS being set to FALSE.\n")
}

```
#Save the data in an RData file 
This is so that the next file can start processing it from there
```{r save_RData}
saveRDS(GSE270199.imputed, paste0(OUTPUT_FILE_LOCATION,"/GSE270199_imputed.rds"))
saveRDS(GSE270199.redmeta, paste0(OUTPUT_FILE_LOCATION,"/GSE270199_redmeta.rds"))
```
