---
title: "MSB1015 - GSE270199 Loading and Cleaning"
author: "Anne Savelkoul - i6326700"
date: "2026-09"
output: html_document

---

#Downloading the data
Download GSE270199_COL3A1_normalizedCounts.txt.gz from https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE270199. Unzip it and place it into the INPUT_FILE_LOCATION you specify in File locations. 

Download the Metadata of GSE270199_COL3A1_normalizedCounts.txt.gz from https://www.ncbi.nlm.nih.gov/Traces/study/?acc=PRJNA1125404&o=acc_s%3Aa. Unzip it and place it into the INPUT_FILE_LOCATION you specify in File locations. 
*RENAME IT TO "GSE270199_SraRunTable.csv"* instead of "SraRunTable.csv" to prevent multiple files being names "SraRunTable.csv".

*Unless specified otherwise, the code attempts to load in the default names of the files. If the names are changed in other ways then adjust the code in Load data.*

#Clear the global environment
  Clear the environment to reduce the chance of issues on reruns
```{r clear_global}
rm(list=ls())
```

#Setting the working directory (wd)
Other files are assumed to be located in the same, or downstream maps from the location of this file. The location of this file is used as a reference point to find the other files. 
```{r working_directory}
#download and load the package for setting the working directory to the current file
if (!require("rstudioapi", quietly = TRUE))
  install.packages("rstudioapi")
library(rstudioapi)  

#Set the wd to the location of this file  
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```
#Library installations and loading 
As I will be using a project (renv) for the MSB1015 project, I check first that the packages that need to be installed are installed.
```{r library_R_file, include=FALSE}
#For cleanliness the checks, installing and loading of the various libraries is in a separate .R file. This file is assumed to be in the same map as this file (in the wd).
source("libraries.R")
```

```{r library_installation}
#Use the function to install the libraries
Fun_install_libraries()
```
```{r library_loading}
#Use the function to load the libraries
Fun_load_libraries()
```
#Load Agi's URF functions
  Adds the functions:
    script_URF
    unsupervisedRF_proximityadd
    doublecentering
```{r}
#The R session aborted very quickly after trying to play with the functions, so I am leaving Agi's functions alone for now.
#source("URF.R")
```

#Load in the checks R file
```{r checks}
#There are some functions to run checks for various things in this separate .R file.
#These are not in the main code as they create quite a bit of bloat that makes the code more difficult to read.
#These are not reusable functions!
source("checks.R")
```
#Load in the custom_functions R file
```{r custom_functions}
#There are some functions to run checks for various things in this separate .R file.
#These are not in the main code to make the main script less bloated and the custom functions easier to find (instead of all of them being spread over the Rmd file).
source("custom_functions.R")
```
#File locations
Set the locations of the files that are loaded in and the locations on where the outputs will go. 
This code will also do a check that the given locations are findable. 
```{r file_locations}
#Make sure to save the wd so that it is easy to use to set file locations
working_directory <- getwd()

#The default INPUT_FILE_LOCATION is in a map where this file is also saved.
INPUT_FILE_LOCATION <- paste0(working_directory,"/INPUTS")
#The default OUTPUT_FILE_LOCATION is in a map where this file is also saved.
OUTPUT_FILE_LOCATION <- paste0(working_directory,"/OUTPUTS")

#-----#

#Make sure that the file locations can be found and in the case of the outputs create a dir if it doesn't exist
Fun_in_out_check()

```
#Settings
Any global settings that are used are set here. 
```{r settings}
#-----#
#In multiple sections:
  #During testing it was often beneficial to run the whole file. Preventing the plots from being printed allows the file to run slightly faster.
  #This setting is mostly used for testing purposes. 
#TAG: add into the settings check
RUN_PLOTS <- TRUE

#In section: remove_missing
  #Missing values are removed from GSE270199, these are the thresholds that are permitted for the gene to be kept. 
#Set the percentage (0 to 100) of zeros that are allowed in the healthy group
  #There are 4 values, setting it to 25 means that 3/4 need to be non-zero
allow_zeros_in_healthy_percent <- 25
#Set the percentage (0 to 100) of zeros that are allowed in the EDS group
  #There are 10 values, setting it to 20 means that 8/10 need to be non-zero
allow_zeros_in_EDS_percent <- 20
#Set which filtering you would like to continue with.
  #Options:
    #"either" --> Either the EDS or the healthy group needs to pass the thresholds for the gene to be kept.
    #"both" --> Both the EDS and the healthy group need to pass the thresholds for the gene to be kept.
    #"difference" --> The difference between the previous two options.
FILTERSETTING <- "either"





#TAG: Add discription of what this setting does
#TAG: add into the settings check
IMPUTE_MISSING <- TRUE
#TAG: Add discription of what this setting does
#TAG: add into the settings check
IMPUTE_ZEROS <- TRUE
#TAG: Add discription of what this setting does
#TAG: turn this into percentages as well if time allows
#TAG: add into the settings check
zeros_for_imputation_EDS = 8
zeros_for_imputation_healthy = 3


#-----#
```
#Settings check
Check if the values that were given in the settings are allowed
```{r}
Fun_settings_check()
```
#Load GSE270199 data
The data is loaded here. 
```{r load_data_GSE270199}
#Check that the required inputs can be found, otherwise stop the code from running
Fun_GSE270199_findability_check()

#-----# 

#Load GSE270199 and fix up the rownames
GSE270199 <- read.delim(paste0(INPUT_FILE_LOCATION,"/GSE270199_COL3A1_normalizedCounts.txt"))
rownames(GSE270199) <- GSE270199[,1]
GSE270199 <- GSE270199[,-1]

#Print a message if GSE270199 exists in the environment (meaning that it was loaded)
if(exists("GSE270199")) cat("GSE270199 has been loaded \n")

#-----# 

#Load the metadata of GSE270199 and fix up the rownames
#I do not remove the Library.Name from the metadata file in case I want to use it for something later
GSE270199.meta <- read.csv(paste0(INPUT_FILE_LOCATION,"/GSE270199_SraRunTable.csv"))
rownames(GSE270199.meta) <- GSE270199.meta[,"Library.Name"]

#Print a message if GSE270199 exists in the environment (meaning that it was loaded)
if(exists("GSE270199.meta")) cat("GSE270199.meta has been loaded \n")

#-----# 

#By default the names of the patients in the normalizedCounts and the metadata do not line up. 
# This is not in the metadata file and needs to be manually adjusted 
# Manually copied the aligning names from https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE270199 and rename the colnames one by one before applying them.

#Get the origninal colnames
GSE270199.colnames <- colnames(GSE270199)

#Replace them with the ones that line up with the metadata
GSE270199.colnames <- gsub(pattern = "C3B_healthy", replacement = "GSM8336700", x = GSE270199.colnames)
GSE270199.colnames <- gsub(pattern = "C2B_healthy", replacement = "GSM8336701", x = GSE270199.colnames)
GSE270199.colnames <- gsub(pattern = "RR_healthy", replacement = "GSM8336702", x = GSE270199.colnames)
GSE270199.colnames <- gsub(pattern = "C1A_healthy", replacement = "GSM8336703", x = GSE270199.colnames)
GSE270199.colnames <- gsub(pattern = "X20.3156_pat", replacement = "GSM8336704", x = GSE270199.colnames)
GSE270199.colnames <- gsub(pattern = "X20.3154_pat", replacement = "GSM8336705", x = GSE270199.colnames)
GSE270199.colnames <- gsub(pattern = "X20.4456_pat", replacement = "GSM8336706", x = GSE270199.colnames)
GSE270199.colnames <- gsub(pattern = "VL_pat", replacement = "GSM8336707", x = GSE270199.colnames)
GSE270199.colnames <- gsub(pattern = "GAU_pat", replacement = "GSM8336708", x = GSE270199.colnames)
GSE270199.colnames <- gsub(pattern = "X19.1597_pat", replacement = "GSM8336709", x = GSE270199.colnames)
GSE270199.colnames <- gsub(pattern = "SF_pat", replacement = "GSM8336710", x = GSE270199.colnames)
GSE270199.colnames <- gsub(pattern = "X22.0547_pat", replacement = "GSM8336711", x = GSE270199.colnames)
GSE270199.colnames <- gsub(pattern = "X22.0548_pat", replacement = "GSM8336712", x = GSE270199.colnames)
GSE270199.colnames <- gsub(pattern = "X20.4222_pat", replacement = "GSM8336713", x = GSE270199.colnames)
	
#Apply them to the colnames
colnames(GSE270199) <- GSE270199.colnames

#Remove GSE270199.colnames from the environment as it is no longer used
rm(GSE270199.colnames)
	
#Check that the people in GSE270199 are also in GSE270199.meta
#Modify this check if time allows so it can be reused for the other data if it gets used.
Fun_GSE270199_meta_check()
```
#Metadata
Check what is available in the metadata and remove un-relevant information / add information that could be useful.
```{r metadata}
#Check metadata

#Break up the output
cat("\n-----\n")
#Create reduced metadata (redmeta) that removes columns that only contain a single thing
keep <- NULL
for (column in 1:ncol(GSE270199.meta)) {
  keep <- c(keep,length(unique(GSE270199.meta[,column])) != 1)
}
GSE270199.redmeta <- GSE270199.meta[,keep]

#Remove ones that are fully unique (cannot check for any batch effects on that anyway)
keep <- NULL
for (column in 1:ncol(GSE270199.redmeta)) {
  keep <- c(keep,length(unique(GSE270199.redmeta[,column])) != nrow(GSE270199.redmeta))
}

GSE270199.redmeta <- GSE270199.redmeta[,keep]

#Manually remove colums that are unrelevant
cat("Remaining columns after automatic removal in GSE270199.redmeta:\n", colnames(GSE270199.redmeta),"\n")
cat("The only one considered interesting is \"disease_state\".\n")
GSE270199.redmeta <- GSE270199.redmeta["disease_state"]

#Break up the output
cat("\n-----\n")

#Rename the values inside the disease_state for ease of use
  #EDS patient --> EDS-HT/JHS
  #healthy control --> healthy
rename <- sub("EDS patient","EDS-HT/JHS", GSE270199.redmeta[,"disease_state"])	
rename <- sub("healthy control","healthy", rename)
GSE270199.redmeta[,"disease_state"] <- rename
cat("The values inside GSE270199.redmeta[\"disease_state\"] have been renamed to align with the data in the metadata files of the other dataset:\n\tEDS patient --> EDS-HT/JHS\n\thealthy control --> healthy\n")

#Break up the output
cat("\n-----\n")
#-----# 

#Remove keep and rename as they are no longer needed
rm(keep, rename, column)

```
#TAG: REMOVE THIS, THIS WAS A TEST ON HOW THE DATA RESPONDED TO GSM8336700 BEING REMOVED
  The answer was confusing, not in a way I want to deal with
```{r temp_test}
#GSE270199 <- GSE270199[colnames(GSE270199) != "GSM8336700"]
#
#temp <- as.data.frame(GSE270199.redmeta[grep("GSM8336700", rownames(GSE270199.redmeta), invert = TRUE),"disease_state"])
#colnames(temp) <- "disease_state"
#rownames(temp) <- rownames(GSE270199.redmeta)[grep("GSM8336700", rownames(GSE270199.redmeta), invert = TRUE)]
#
#GSE270199.redmeta <- temp
```
#Missing values
Check if the loaded data has any missing values. 
```{r check_missing_1}
#Break up the output
cat("\n-----\n")

#TAG: turn this into a reusable function and place it in the checks file
#Check if any values are NA 
if (!(NA %in% GSE270199)){
  cat("There are no \"NA\" values in GSE270199 \n")
} else if (NA %in% GSE270199){
  warning("There are \"NA\" values in GSE270199 \n")
} else {
  warning("Something unknown went wrong\n")
}

#TAG: turn this into a reusable function and place it in the checks file
#Check how many zeros are in the data
cat(paste0("There are ", sum(GSE270199 == 0), " zero values in GSE270199. \n",
           "That is ", 
           round((sum(GSE270199 == 0)/(ncol(GSE270199) * nrow(GSE270199)))*100, 2), 
           "% of the data.\n"))

#TAG: turn this into a reusable function and place it in the checks
#Check what the smallest value is
cat(paste0("The smallest value in GSE270199 (ignorming zeros) is ", min(GSE270199[GSE270199 != 0]),"."))

#Break up the output
cat("\n-----\n")

if (!(NA %in% GSE270199.meta)){
  cat("There are no \"NA\" values in GSE270199.meta. \n")
} else if (NA %in% GSE270199.meta){
  warning("There are \"NA\" values in GSE270199.meta. \n")
} else {
  warning("Something unknown went wrong\n")
}

#Break up the output
cat("\n-----\n")
```
#GSE270199 missing data processing
GSE270199 has many zeros, so those values need to be processed.
```{r remove_missing_GSE270199}
#Get the patient identifiers so that the zero removal can be done while keeping the groups in mind.
healthy_group <- rownames(GSE270199.redmeta)[grep("healthy", GSE270199.redmeta[,"disease_state"])]
EDS_group <- rownames(GSE270199.redmeta)[grep("EDS-HT", GSE270199.redmeta[,"disease_state"])]

#Convert the percentage set in the settings to a number that is used for the comparison
allow_zeros_healthy <- (length(healthy_group)*(allow_zeros_in_healthy_percent/100))

#Convert the percentage to a number that is used for the comparison
allow_zeros_EDS <- (length(EDS_group)*(allow_zeros_in_EDS_percent/100))

#Loop over GSE270199 to find the ones that have too many zeros
keep_either <- NULL
keep_both <- NULL
for (i in 1:nrow(GSE270199)){
  keep_EDS <- (sum(GSE270199[i,EDS_group] == 0) <= allow_zeros_EDS)
  keep_healthy <- (sum(GSE270199[i,healthy_group] == 0) <= allow_zeros_healthy)
  
  #It can be decided to keep 
    #Either: Either one or the other group needs to pass their threshold.
    #Both: Both of the groups need to pass their threshold.
  keep_either <- c(keep_either, keep_EDS | keep_healthy)
  keep_both <- c(keep_both, keep_EDS & keep_healthy)
}

#Get the filtered dataframes based on the keep's that were made
#Filtered of the dataframe if either of the groups pass the threshold
GSE270199.filtered_either <- GSE270199[keep_either,]
#Filtered of the dataframe if both of the groups pass the threshold
GSE270199.filtered_both <- GSE270199[keep_both,]
#Filtered of the dataframe if only one of the groups pass the threshold
GSE270199.filtered_difference <- GSE270199[(!keep_both & keep_either),]


#Remove the temp objects from the environment
rm(keep_either, keep_both, keep_EDS, keep_healthy, i)

```
#Select filtering using missingness plots
Plot all of the different data created in the script so far to compare the different types of data. 
```{r missingness_plots}
#Create color_groups to color the labels of the samples so it is easier to see what the controls are vs the patients
color_groups <- ifelse(GSE270199.redmeta[colnames(GSE270199),] == "EDS-HT/JHS", "#FF0000", "#0000FF")

#Plot the missingness plots
if (RUN_PLOTS){
  Fun_missingness(GSE270199, color_groups, "GSE270199")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_missingness(GSE270199.filtered_either, color_groups, "GSE270199 filtered, minimum one passes")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_missingness(GSE270199.filtered_both, color_groups, "GSE270199 filtered, both pass")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_missingness(GSE270199.filtered_difference, color_groups, "GSE270199 difference between min one and both")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}

```
#Compare the pca plots of the different filtered data
Create the PCA objects
```{r pca_plots_1_1}
#Calculate the PCA for the non-filtered data
PCA.original <- pca(object = t(as.matrix(GSE270199)), 
                   method = "svd", nPcs = 9, center = TRUE)
#Calculate the PCA for filtered either
PCA.filtered_either <- pca(object = t(as.matrix(GSE270199.filtered_either)), 
                   method = "svd", nPcs = 9, center = TRUE)
#Calculate the PCA for filtered both
PCA.filtered_both <- pca(object = t(as.matrix(GSE270199.filtered_both)), 
                   method = "svd", nPcs = 9, center = TRUE)
#Calculate the PCA for the difference
PCA.filtered_difference <- pca(object = t(as.matrix(GSE270199.filtered_difference)), 
                   method = "svd", nPcs = 9, center = TRUE)
```
Visualize the scores of the first two PC's
```{r pca_plots_1_2}
if (RUN_PLOTS){
  Fun_plot_PCA_scores(PCA = PCA.original, 
                      metadata = GSE270199.redmeta, 
                      firstPC = "PC1", 
                      secondPC = "PC2", 
                      title = "Original PCA scores")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_plot_PCA_scores(PCA = PCA.filtered_either, 
                      metadata = GSE270199.redmeta, 
                      firstPC = "PC1", 
                      secondPC = "PC2", 
                      title = "PCA scores, either filtered")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_plot_PCA_scores(PCA = PCA.filtered_both, 
                      metadata = GSE270199.redmeta, 
                      firstPC = "PC1", 
                      secondPC = "PC2", 
                      title = "PCA scores, both filtered")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_plot_PCA_scores(PCA = PCA.filtered_difference, 
                      metadata = GSE270199.redmeta, 
                      firstPC = "PC1", 
                      secondPC = "PC2", 
                      title = "PCA scores difference between either and both filters")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
```
Visualize the loadings of the first two PC's
```{r pca_plots_1_3}
if (RUN_PLOTS){
  Fun_plot_PCA_loadings(PCA = PCA.original, 
                        firstPC = "PC1", 
                        secondPC = "PC2", 
                        title = "Original PCA scores")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_plot_PCA_loadings(PCA = PCA.filtered_either, 
                        firstPC = "PC1", 
                        secondPC = "PC2", 
                        title = "PCA scores, either filtered")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_plot_PCA_loadings(PCA = PCA.filtered_both, 
                        firstPC = "PC1", 
                        secondPC = "PC2", 
                        title = "PCA scores, both filtered")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_plot_PCA_loadings(PCA = PCA.filtered_difference, 
                        firstPC = "PC1", 
                        secondPC = "PC2", 
                        title = "PCA scores difference between either and both filters")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
```
#Biplots
Check the top most influential genes using a biplot
```{r biplots}
if (RUN_PLOTS){
  Fun_plot_biplot(PCA = PCA.original, 
                  top = 50, 
                  firstPC = "PC1", 
                  secondPC = "PC2", 
                  title = "Biplot using the top 50 genes for the non-processed data.")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_plot_biplot(PCA = PCA.filtered_either, 
                  top = 50, 
                  firstPC = "PC1", 
                  secondPC = "PC2", 
                  title = "Biplot using the top 50 genes for the either filtering.")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_plot_biplot(PCA = PCA.filtered_both, 
                  top = 50, 
                  firstPC = "PC1", 
                  secondPC = "PC2", 
                  title = "Biplot using the top 50 genes for the both filtering.")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_plot_biplot(PCA = PCA.filtered_difference, 
                  top = 50, 
                  firstPC = "PC1", 
                  secondPC = "PC2", 
                  title = "Biplot using the top 50 genes for the difference filtering.")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}

```
#Select what filtered data to continue forward with
```{r select_filtered}
#Depending on the setting specified, continue on with one of the filtered sets
if (FILTERSETTING == "either"){
  GSE270199.filtered <- GSE270199.filtered_either
  cat("Using genes that pass the threshold in either of the groups.\n")
} else if (FILTERSETTING == "both"){
  GSE270199.filtered <- GSE270199.filtered_both
  cat("Using genes that pass the threshold in both groups.\n")
} else if (FILTERSETTING == "difference"){
  GSE270199.filtered <- GSE270199.filtered_difference
  cat("Using the difference.\n")
} else {
  warning("Something unknown went wrong, stopping the code.\n")
  stop()
}
```
#Data imputation
Impute the remaining zeros within their respective groups depending on the setting provided.
```{r imputation}
if (IMPUTE_MISSING & !IMPUTE_ZEROS){
  #IMPUTE_MISSING --> Impute the missing values where possible
  #!IMPUTE_ZEROS --> Do not impute to zeros
  GSE270199.imputed <- Fun_impute_data(GSE270199.filtered, 
               allow_zeros_EDS = allow_zeros_EDS, 
               allow_zeros_healthy = allow_zeros_healthy,
               IMPUTE_TO_ZERO = FALSE
               )
  cat("Imputed missing where possible.\n")
  
} else if (IMPUTE_MISSING & IMPUTE_ZEROS){
  #IMPUTE_MISSING --> Impute the missing values where possible
  #IMPUTE_ZEROS --> Also impute to zero if the thresholds for zeros_for_imputation_EDS / zeros_for_imputation_healthy specified in settings is passed
  GSE270199.imputed <- Fun_impute_data(GSE270199.filtered, 
               allow_zeros_EDS = allow_zeros_EDS, 
               allow_zeros_healthy = allow_zeros_healthy,
               IMPUTE_TO_ZERO = TRUE,
               zeros_for_imputation_EDS = zeros_for_imputation_EDS,
               zeros_for_imputation_healthy = zeros_for_imputation_healthy
               )
  cat("Imputed missing and to zeros where possible.\n")
  
} else if (!IMPUTE_MISSING){
  #!IMPUTE_MISSING --> Do not impute anything
    #Still set it to the imputed variable name so that the following code can run it
  GSE270199.imputed <- GSE270199.filtered
  cat("Did not impute anything.\n")
  
} else {
  warning("Something unknown went wrong, stopping the code.\n")
  stop()
}
```
#Compare before and after imputation
Compare what the imputation has done to the missingness of the data.
```{r imputation_result}
#DUPLICATE: color_groups is created above as well, but repeated for clarity
#Create color_groups to color the labels of the samples so it is easier to see what the controls are vs the patients
color_groups <- ifelse(GSE270199.redmeta[colnames(GSE270199),] == "EDS-HT/JHS", "#FF0000", "#0000FF")

#Visualize the missingness plots
  #NOTE: The implementation of the color_groups causes a warning. 
if (RUN_PLOTS){
  Fun_missingness(GSE270199.filtered, color_groups, "Before imputation")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_missingness(GSE270199.imputed, color_groups, "After imputation")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
```
#Missing values after correction
Check the progress of the missing value removal. 
```{r check_missing_2}
#Check how many zeros are in the data
cat(paste0("There are ", sum(GSE270199.filtered == 0), " zero values in GSE270199.filtered before imputation \n",
           "That is ", 
           round((sum(GSE270199.filtered == 0)/(ncol(GSE270199.filtered) * nrow(GSE270199.filtered)))*100, 2), 
           "% of the data.\n"))
#Check how many zeros are in the data
cat(paste0("There are ", sum(GSE270199.imputed == 0), " zero values in GSE270199.imputed after imputation \n",
           "That is ", 
           round((sum(GSE270199.imputed == 0)/(ncol(GSE270199.imputed) * nrow(GSE270199.imputed)))*100, 2), 
           "% of the data.\n"))
```
#Compare the pca's before and after imputation
Calculate the PCA's
```{r pca_plots_2_1}
#Calculate the PCA for the chosen filtered data
PCA.filtered <- pca(object = t(as.matrix(GSE270199.filtered)), 
                   method = "svd", nPcs = 9, center = TRUE)
#Calculate the PCA for the chosen imputed data
PCA.imputed <- pca(object = t(as.matrix(GSE270199.imputed)), 
                   method = "svd", nPcs = 9, center = TRUE)
```
Visualize the scores of the first two PCs
```{r pca_plots_2_2}
if (RUN_PLOTS){
  Fun_plot_PCA_scores(PCA = PCA.filtered, 
                      metadata = GSE270199.redmeta, 
                      firstPC = "PC1", 
                      secondPC = "PC2", 
                      title = "PCA scores before imputation")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_plot_PCA_scores(PCA = PCA.imputed, 
                      metadata = GSE270199.redmeta, 
                      firstPC = "PC1", 
                      secondPC = "PC2", 
                      title = "PCA scores after imputation")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
```
Visualize the loadings of the first two PCs
```{r pca_plots_2_3}
if (RUN_PLOTS){
  Fun_plot_PCA_loadings(PCA = PCA.filtered, 
                        firstPC = "PC1", 
                        secondPC = "PC2", 
                        title = "PCA loadings before imputation")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_plot_PCA_loadings(PCA = PCA.imputed, 
                        firstPC = "PC1", 
                        secondPC = "PC2", 
                        title = "PCA loadings after imputation")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
```
#Biplots
Compare how the biplot has changed due to the imputation
```{r biplots}
if (RUN_PLOTS){
  Fun_plot_biplot(PCA = PCA.filtered, 
                  top = 50, 
                  firstPC = "PC1", 
                  secondPC = "PC2", 
                  title = "Biplot using the top 50 genes for the filtered data.")
} else {
  cat("Plots not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_plot_biplot(PCA = PCA.imputed, 
                  top = 50, 
                  firstPC = "PC1", 
                  secondPC = "PC2", 
                  title = "Biplot using the top 50 genes for the imputed data.")
} else {
  cat("Plots not visualized due to RUN_PLOTS being set to FALSE.\n")
}

```
#Save the data in an RData file 
This is so that the next file can start processing it from there
```{r save_RData}
#TAG: DO THIS INSTEAD OF HAVING EVERYTHING IN ONE FILE'
saveRDS(GSE270199.imputed, paste0(OUTPUT_FILE_LOCATION,"/GSE270199_imputed.rds"))
saveRDS(GSE270199.redmeta, paste0(OUTPUT_FILE_LOCATION,"/GSE270199_redmeta.rds"))
```

#TAG: Put what is below this into a different file:
--> MSB1015-GSE270199_Processing.Rmd


#PCA on only disease
```{r PCA_on_disease}
#Subset to only get the patient data
GSE270199.disease_only <- GSE270199.imputed[EDS_group]

#Run a PCA on the patient only data
PCA.disease_only <- pca(object = t(as.matrix(GSE270199.disease_only)), 
                   method = "svd", nPcs = 9, center = TRUE)


if (RUN_PLOTS){
  Fun_plot_PCA_scores(PCA = PCA.disease_only, 
                      metadata = GSE270199.redmeta, 
                      firstPC = "PC1", 
                      secondPC = "PC2", 
                      title = "PCA scores (PC1 vs PC2) of just the patients")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}

if (RUN_PLOTS){
  Fun_plot_PCA_scores(PCA = PCA.disease_only, 
                        firstPC = "PC2", 
                        secondPC = "PC3", 
                        title = "PCA scores (PC2 vs PC3) of just the patients")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_plot_PCA_scores(PCA = PCA.disease_only, 
                        firstPC = "PC3", 
                        secondPC = "PC4", 
                        title = "PCA scores (PC3 vs PC4) of just the patients")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}
if (RUN_PLOTS){
  Fun_plot_PCA_scores(PCA = PCA.disease_only, 
                        firstPC = "PC4", 
                        secondPC = "PC5", 
                        title = "PCA scores (PC4 vs PC5) of just the patients")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}




if (RUN_PLOTS){
  Fun_plot_PCA_loadings(PCA = PCA.disease_only, 
                        firstPC = "PC1", 
                        secondPC = "PC2", 
                        title = "PCA loadings (PC1 vs PC2) of just the patients")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}

if (RUN_PLOTS){  
  Fun_plot_biplot(PCA = PCA.disease_only, 
                  top = 50, 
                  firstPC = "PC1", 
                  secondPC = "PC2", 
                  title = "Biplot using the top 50 genes for the just patient data.")
} else {
  cat("Plot not visualized due to RUN_PLOTS being set to FALSE.\n")
}

```

#WIP
```{r}
#Get the top loadings for both PC1 and PC2 to use as the background network
top_loadings <- as.data.frame(PCA.disease_only@loadings) %>%
    mutate(
      contribution = sqrt(eval(parse(text="PC1"))^2 + eval(parse(text="PC2"))^2)
    ) %>%
    arrange(desc(contribution)) %>%
    head(round(length(rownames(as.data.frame(PCA.disease_only@loadings)))*0.75))
top_loadings <- rownames(top_loadings)

#Get loadings that pass 0.03 for PC1
#temp <- as.data.frame(PCA.disease_only@loadings)["PC1"]
#top_loadings_PC1 <- rownames(temp)[abs(temp) > 0.01]

#Get loadings that pass 0.03 for PC2
#temp <- as.data.frame(PCA.disease_only@loadings)["PC2"]
#top_loadings_PC2 <- rownames(temp)[abs(temp) > 0.01]

#Get the top loadings 
#top_loadings <- unique(top_loadings_PC1, top_loadings_PC2)

#Check that Cytoscape can be reached
cytoscapePing()

#If this section of code is re-run then the code runs into issues, to prevent this, remove if there is a networkSuid in the environment already and reload it
if (exists("networkSuid")){
  cat("networkSuid was already made, removing it and remaking it to prevent issues with the code rerunning.\n")
  RCy3::deleteNetwork(networkSuid)
  rm(networkSuid)
  cat("networkSuid has been deleted and will be remade.\n")
} else if (!exists("networkSuid")){
  cat("networkSuid does not exist in the environment, creating new network. \n")
}


#Build the command to send to Cytoscape
cmd <- paste0(
  'string protein query ',
  'query="', top_loadings, '" ',
  'networkType=full STRING network ',
  'cutoff=0.4 ',
  'newNetName=Base Network ',
  'species="Homo sapiens" ',
  'limit=100 ',
  'showFlatNodeDesign=true ',
  'showSTRINGstyleLabels=false ',
  'showStructureImages=false'
  )

#Send the command to Cytoscape to build the network
RCy3::commandsRun(cmd)

#Get the network's SUID
networkSuid = getNetworkSuid()



#-----#

```


#WIP AI CODE
```{r WIP_AI_code}
if(FALSE){
  #Skip this code for now
  
# 2. Extract loadings (rotation matrix), convert to data frame, and select PC1
loadings_df <- PCA.imputed@loadings %>%
  as.data.frame() %>%
  rownames_to_column(var = "Variable") %>%
  select(Variable, PC1)

# 3. Find the top 20 most important variables based on absolute loading value for PC1
top_20_loadings <- loadings_df %>%
  mutate(Abs_PC1 = abs(PC1)) %>%
  arrange(desc(Abs_PC1)) %>%
  head(20) %>%
  # Reorder factor levels for proper plotting order
  mutate(Variable = reorder(Variable, PC1))


if (RUN_PLOTS){
  # 4. Plot the top 20 loadings using ggplot2
  ggplot(top_20_loadings, aes(x = Variable, y = PC1, fill = PC1 > 0)) +
    geom_col(show.legend = FALSE) +
    coord_flip() + # Makes horizontal bar chart for readability
    theme_minimal() +
    labs(
      title = "Top 20 Most Important Loadings for PC1",
      x = "Original Variables",
      y = "Loading Score (PC1)"
    ) +
    scale_fill_manual(values = c("TRUE" = "#2b8cbe", "FALSE" = "#e34a33"))
} else {
  cat("Plots not visualized due to RUN_PLOTS being set to FALSE.\n")
}


top_20_loadings$Variable
}
```

#Top 20:
ENSG00000164530 ENSG00000111057 ENSG00000110723 ENSG00000060718 ENSG00000167641 ENSG00000135480 ENSG00000267368 ENSG00000148677 ENSG00000243137 ENSG00000165072 ENSG00000250337 ENSG00000164176 ENSG00000170345 ENSG00000021645 ENSG00000064787 ENSG00000272145 ENSG00000143473 ENSG00000205426 ENSG00000064205 ENSG00000196460
#Top 50:
ENSG00000164530 ENSG00000111057 ENSG00000110723 ENSG00000060718 ENSG00000167641 ENSG00000135480 ENSG00000267368 ENSG00000148677 ENSG00000243137 ENSG00000165072 ENSG00000250337 ENSG00000164176 ENSG00000170345 ENSG00000021645 ENSG00000064787 ENSG00000272145 ENSG00000143473 ENSG00000205426 ENSG00000064205 ENSG00000196460 ENSG00000246223 ENSG00000145242 ENSG00000129682 ENSG00000267390 ENSG00000111087 ENSG00000153707 ENSG00000244649 ENSG00000248576 ENSG00000133110 ENSG00000196616 ENSG00000248187 ENSG00000165617 ENSG00000123689 ENSG00000117115 ENSG00000241644 ENSG00000282860 ENSG00000257894 ENSG00000145681 ENSG00000117643 ENSG00000002933 ENSG00000155011 ENSG00000177459 ENSG00000157303 ENSG00000176406 ENSG00000188393 ENSG00000170558 ENSG00000137331 ENSG00000055732 ENSG00000281501 ENSG00000178776



#Calculate log2FC values
The formula used to calculate the fold change was found at:
https://www.biostars.org/p/342756/
```{r log2FC}
#Change the whole dataset to log2 notation
GSE270199.imputed.log2 <- log2(GSE270199.imputed)
#Remove the -Inf's created from taking the log2 of zero by setting it to zero
GSE270199.imputed.log2[GSE270199.imputed.log2 == -Inf] <- 0

#DUPLICATE: These groups were specified before but are here again for clarity
#Get the identifiers of the two groups
healthy_group <- rownames(GSE270199.redmeta)[grep("healthy", GSE270199.redmeta[,"disease_state"])]
EDS_group <- rownames(GSE270199.redmeta)[grep("EDS-HT", GSE270199.redmeta[,"disease_state"])]

#Pre-make the dataframe to store the data in
log2FC_dataframe <- data.frame(row.names = rownames(GSE270199.imputed.log2))
#Pre-allocate to save the top genes
top_log2FC_list <- list()
#Pre-allocate to save the differential ones
diff_log2FC_list <- list()

#Loop to get the log2FC's of the patients vs the controls, one-on-one
for (control in 1:length(healthy_group)){
  #cat for some feedback
  cat(paste0("C",control," "))
  for (patient in 1:length(EDS_group)){
    #cat for some feedback
    cat(paste0("P",patient," "))
    
    #Get a run_name to save the name of the log2FC
    run_name <- paste0(EDS_group[patient],"_",healthy_group[control])
    
    #Calculate the log2FC using subtraction:
      #log2(A)-log2(B) == log2(A/B)
    temp <- GSE270199.imputed.log2[,EDS_group[patient]] - GSE270199.imputed.log2[,healthy_group[control]]
    #Set the temp to the run_name
    log2FC_dataframe[[run_name]] <- temp
    
    #Get the differential genes that are used to build the STRING network 
      #This makes the assumption that this prevents loading in housekeeping genes that do not change anyway
      #An abs value of 1 or above means that the gene at least doubled or halved it's expression
    diff_log2FC_list[[run_name]] <- rownames(log2FC_dataframe[(abs(temp) >= 1),])
    
    #Get the top genes for this comparison
    temp_top <- temp %>% 
                  as.data.frame(row.names = rownames(log2FC_dataframe)) %>%
                  slice_max(abs(temp), n = 20) %>% 
                  rownames(temp)
    #Set the top_temp to the run_name
    top_log2FC_list[[run_name]] <- temp_top
      
  }
}

#Extract the genes for the background network
STRING_background_network <- unique(unlist(diff_log2FC_list))
#Set it to something that can be queried
STRING_background_network <- readr::format_csv(as.data.frame(head(STRING_background_network,250)), 
                                               col_names=F, escape = "double", eol =",")

#Extract the genes that came out on top at least once
STRING_top_node_genes <- unique(unlist(top_log2FC_list))
#Set it to something that can be queried
STRING_top_node_genes <- readr::format_csv(as.data.frame(STRING_top_node_genes), 
                                           col_names=F, escape = "double", eol =",")

#Remove the temp objects from the environment
rm(temp, temp_top, patient, control, run_name)

```

#Load network into Cytoscape and Diffusion
These two tasks are in the same block so that rerunning the block is possible 
  It reloads the Cytoscape network each time, to prevent running into errors
Use Cytoscape to run the diffusion on the network and extract the heat and rank information.

```{r load_network_and_diffusion, echo=TRUE, output=TRUE, include=TRUE}
#See if Cytoscape can be reached
  #This automatically stops the code when Cytoscape cannot be reached (built into the function)
cytoscapePing()


#If this section of code is re-run then the code runs into issues, to prevent this, remove if there is a networkSuid in the environment already and reload it
if (exists("networkSuid")){
  cat("networkSuid was already made, removing it and remaking it to prevent issues with the code rerunning.\n")
  RCy3::deleteNetwork(networkSuid)
  rm(networkSuid)
  cat("networkSuid has been deleted and will be remade.\n")
} else if (!exists("networkSuid")){
  cat("networkSuid does not exist in the environment, creating new network. \n")
}


#Build the command to send to Cytoscape
cmd <- paste0(
  'string protein query ',
  'query="', STRING_background_network, '" ',
  'networkType=physical subnetwork ',
  'cutoff=0.4 ',
  'newNetName=Base Network ',
  'species="Homo sapiens" ',
  'limit=100 ',
  'showFlatNodeDesign=true ',
  'showSTRINGstyleLabels=false ',
  'showStructureImages=false'
  )

#Send the command to Cytoscape to build the network
RCy3::commandsRun(cmd)

#Get the network's SUID
networkSuid = getNetworkSuid()



#-----#


#Get pre-allocated object to save the diffusion output that can be merged into
  #display name and query term are loaded in so that the merge command can merge the dataframes using that as a reference
diffusion_output <- getTableColumns(
    table = "node",
    columns = c("display name","query term")
)

#TAG: Remove the limitation (was added for testing the code)
for (run_name in colnames(log2FC_dataframe)[c(1,2)]){
  
  #Select the source that the diffusion starts from 
  gene_selection <- c(rownames(log2FC_dataframe)[abs(log2FC_dataframe[run_name]) >= 2])
  
  #Select nodes in Cytoscape network
  selectNodes(
      gene_selection,
      by.col = "query term",
      preserve.current.selection = FALSE
  )
  
  #Setting the time of the diffusion command any lower than the maximum 1 disperses no heat at all
  #Neither does using the diffusion diffuse command
    #I assume there is a bug in the diffusion algorithm of Cytoscape that I am unqualified to do anything about
  #TAG: CHANGE THE TIME
  RCy3::commandsRun('diffusion diffuse_advanced time=0.1')
  
  
  #Rename the diffusion output so that the results don't get mixed up
  renameTableColumn(
      table = "node",
      column = "diffusion_output_heat",
      new.name = paste0(run_name, "_heat")
  )
  renameTableColumn(
      table = "node",
      column = "diffusion_output_rank",
      new.name = paste0(run_name, "_rank")
  )
  
  #Get the diffusion results from Cytoscape
  temp_output <- getTableColumns(
      table = "node",
      columns = c("display name","query term", paste0(run_name, "_heat"), paste0(run_name, "_rank"))
  )
  
  #Round the output to both filter out extremely low heat and make downstream easier
  temp_output[paste0(run_name, "_heat")] <- round(temp_output[paste0(run_name, "_heat")], 3)
  

  #Merge the new result to the diffusion_output
  diffusion_output <- merge(diffusion_output, temp_output)
}

#Remove the temp objects
#rm(temp_output, run_name, gene_selection)

head(diffusion_output)
```



#WIP - Testing comparisons
```{r}
#pearson
diff1 <- diffusion_output$GSM8336713_GSM8336703_heat
diff2 <- diffusion_output$GSM8336712_GSM8336703_heat
cat("Pearson correlation:\t", cor(diff1, diff2, method = "pearson"), "\n")

#spearman
cat("Spearman correlation:\t", cor(diff1, diff2, method = "spearman"), "\n")

#kendall
cat("Kendall correlation:\t", cor(diff1, diff2, method = "kendall"), "\n")

#euclidean
cat("Euclidian distance:\t", sqrt(sum((diff1 - diff2)^2)), "\n")

#euclidean normalized
d1 <- diff1 / sum(diff1)
d2 <- diff2 / sum(diff2)
cat("Normalized Euclidian distance:\t", sqrt(sum((d1 - d2)^2)), "\n")


#cosine
cat("Cosine:\t", 
    sum(diff1 * diff2) / (sqrt(sum(diff1^2)) * sqrt(sum(diff2^2))), 
    "\n")


#Difference
results <- as.data.frame(matrix(nrow = length(diff1), ncol = 0))
results$difference <- diff1 - diff2
results$abs_difference <- abs(diff1 - diff2)
#results[order(-results$abs_difference), ]

#Make a comparison bit use and offset so that nodes that are very similar are also considered equal
results$direction <- ifelse(diff1 > diff2+0.05,
  "Diffusion 1",
  ifelse(diff1 < diff2-0.05,
    "Diffusion 2",
    ifelse(diff1 == diff2 & diff1 == 0,
           "No heat",
           ifelse(diff1 == 0 & diff2 != 0,
             "Only Diffusion 2",
             ifelse(diff1 != 0 & diff2 == 0,
                "Only Diffusion 1",
                 ifelse(diff1 <= diff2+0.05 & diff1 >= diff2-0.05,
                    "Equal",
                    "Something went wrong"))))))


#Also make a comparison encoded as colors for easier passthrough mapping in Cytoscape
results$direction_colors <- ifelse(diff1 > diff2+0.05,
  "red",
  ifelse(diff1 < diff2-0.05,
    "blue",
    ifelse(diff1 == diff2 & diff1 == 0,
           "white",
           ifelse(diff1 == 0 & diff2 != 0,
             "lightblue",
             ifelse(diff1 != 0 & diff2 == 0,
                "pink",
                 ifelse(diff1 <= diff2+0.05 & diff1 >= diff2-0.05,
                    "lightgreen",
                    "Something went wrong"))))))
              
cat(paste0(sum(results["direction"]=="No heat"), 
    " nodes (",
    round((sum(results["direction"]=="No heat")/nrow(results["direction"]))*100, 2),
    "%) carry no heat in both the comparisons.\n"))

cat(paste0(sum(results["direction"]=="Equal"), 
    " nodes (",
    round((sum(results["direction"]=="Equal")/nrow(results["direction"]))*100, 2),
    "%) are equal between the comparisons.\n"))

cat(paste0(sum(results["direction"]=="Diffusion 1"), 
    " nodes (",
    round((sum(results["direction"]=="Diffusion 1")/nrow(results["direction"]))*100, 2),
    "%) carry more heat in the first diffusion.\n"))

cat(paste0(sum(results["direction"]=="Only Diffusion 1"), 
    " nodes (",
    round((sum(results["direction"]=="Only Diffusion 1")/nrow(results["direction"]))*100, 2),
    "%) only have heat in diffusion 1.\n"))

cat(paste0(sum(results["direction"]=="Diffusion 2"), 
    " nodes (",
    round((sum(results["direction"]=="Diffusion 2")/nrow(results["direction"]))*100, 2),
    "%) nodes carry more heat in the second diffusion.\n"))

cat(paste0(sum(results["direction"]=="Only Diffusion 2"), 
    " nodes (",
    round((sum(results["direction"]=="Only Diffusion 2")/nrow(results["direction"]))*100, 2),
    "%) only have heat in diffusion 2.\n"))

diffusion <- results["direction_colors"]
rownames(diffusion) <- diffusion_output$"display name"

loadTableData(
    diffusion,
    data.key.column = "row.names",
    table.key.column = "display name"
)


```



#Get scores dataframe
#TAG: Probably not useful
```{r}
scores_dataframe <- log2FC_dataframe
scores_dataframe[abs(scores_dataframe) > 3] <- 3
scores_dataframe[abs(scores_dataframe) > 2.5 & abs(scores_dataframe) < 3] <- 2.5
scores_dataframe[abs(scores_dataframe) > 2 & abs(scores_dataframe) < 2.5] <- 2
scores_dataframe[abs(scores_dataframe) > 1.5 & abs(scores_dataframe) < 2] <- 1.5
scores_dataframe[abs(scores_dataframe) > 1 & abs(scores_dataframe) < 1.5] <- 1
scores_dataframe[abs(scores_dataframe) < 1] <- 0

colnames(scores_dataframe)

rownames(scores_dataframe)[scores_dataframe["GSM8336713_GSM8336703"] >= 2]
```

#Think about the class imbalance 
#Get this pipeline started

#Continuation ideas:
Get differential expression between healthy and diseased (for both of the datasets)
Look into that heat dispersion algorithm that was covered in Network Biology
  Instead of looking what is different, look at what system/pathway is different
Compare the differences on that level
